Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)

### 2024-12-03 responseToReviews.md

Page 8 figure 5. The extrapolation to zero amplitude method should be attributed to Cambell and Maffei (J Physiol) 1970). Campbell and Kulikowski (J Physiol, 1972) made a detailed comparison to psychophysics.

> We have now included these references

Figure 5 is an oversimplification. As it doesn’t take into account how the additive noise mixes with the internal VEP response.  This was modeled in Norcia et al., (Vision Res, 1989). The additive noise adds a bias to the measurement, especially if the slope of the response function is shallow. The authors should try to connect the linear regression approach for threshold measurement to the sigmoidal non-linearity, as they are different models of the CRF. In practice the difference at the low contrast end is probably too small to matter and the saturation region is excluded.

> We have now noted that it is important to consider noise for situations where the SNR is low or the slope is shallow and we expand on the point that you have to stick to the part of the CRF where the curve is more or less linear - as we showed explicitly in the figure. The code also includes an explicit 'noise baseline' that can be used to demonstrate changes in the intercept as a function of SNR.

@@

Page 12, better to state whether attention effects were consistent with response gain or contrast gain to make the result more concrete for the reader.

> Reynolds and Heeger (and Boynton) point out that the range of different types of change reported in the literature (response, contrast or mixtures of both) is explained by subtle differences in stimulus configuration (for example, the size and location of the 'probe' and any masking components relative to the size of the 'attentional spotlight'). We now note that the underlying model is one of contrast gain control (multiplicative amplification of the input rather than the output).

Discussion of time varying response paragraph on page 12 is not that relevant to contrast response function measurement, nor is the BCI section.  These applications of SSVEP and could be deleted or the relationship to contrast response function made clearer.

> Originally, we envisaged this document as a primer on SSVEP contrast measurements for those coming to the field for the first time. In particular, we were struck by the large number of recent papers that use multi-frequency SSVEP contrast flicker as an input to BCI systems. There have been at least 400 such publications ('SSVEP AND BCI') in the last 5 years alone (compared to around 10 for amblyopia ('SSVEP AND amblyopia') and 80 for ('SSVEP AND clinical'). Some of these publications measure dynamic changes in covert spatial attention, some measure changes in featural attention, some measure fixation. We know from personal experience that many of the researchers using contrast flicker as a BCI input would benefit from a slightly more applied guide such as this and the BCI section adds very little to the overall length of the manuscipt at this point.

@@

Page, 14. It’s not clear from the text how one would study feedback with the SSVEP, please elaborate.

> In the paper we write that we would look for modulation of subcortical responses due to top-down signals. The most natural things to look for might be interocular normalization (because eye layers are segregated in the LGN) and attentional gain control (because covert spatial attentional modulation must arise in cortex). This is the type of feedback we are thinking of - not cortico-cortico feedback (which is hard to isolate with SSVEP).

@@

> This is a good point and we now include the Boyd reference in the section on fingerprinting.

## Reviewer: 2

# Comments to the Author
This manuscript is intended to be a review of the field of steady-state visual evoked potentials (ssVEPs) over the past 70 years.  The authors state in the introductory paragraph “Here we will describe how an EEG method known as the steady state visually evoked potential (SSVEP) technique has contributed to our understanding of human contrast processing in health, disease and throughout development.” The material included and the references cited, however, are quite limited and omit seminal work in this area.

> We thank the reviewer for their helpful comments. As we noted above, our intention here was to provide an introduction to the SSVEP technique with appropriate simple code samples that would allow relative newcomers to the field to implement the technique. We are, however, keen to avoid misrepresenting the history of SSVEP research and are happy to include appropriate references.

@@

1. All abbreviations should be defined upon first use.

> Done

2. The following statement is unclear and inaccurate: “On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on.” The term “on-off flicker” is typically used for simple luminance flicker without spatial pattern contrast. In the context of pattern stimulation, “on-off” can refer to “appearance-disappearance” of a spatial pattern with its contrast modulated in time. When the pattern appears and disappears into a uniform field of equal space-average luminance, the light and dark elements are modulated in counterphase. Any linear contribution to the response elicited by each set of elements (e.g., bars or checks) is expected to be canceled when summed at the electrode site on the scalp. Thus, the response at 1F to this stimulus is in fact a nonlinear response.

> Thank you for this explanation. We have now reworded the paper to make this clearer. We did take care to define what we mean by 'on-off' flicker in the preceding sentence but we now clarify the alternative 'appearence/disappearence' terminology. The point we were making was that odd harmonic responses (1F) are a feature of this type of modulation because the responses to the appearance and disappearance of a pattern (or mean field) need not be completely symmetrical (see Zemon, et al 1988 for an explicit demonstration).

@@

> We have included three earlier papers by Regan on this topic.

4. The statement “This normalization, achieved through a computation called ‘contrast gain control’” again neglects to cite the original work that defined “contrast gain control” and the authors ignored a nonlinear biophysical model proposed to explain this phenomenon in ssVEP contrast response functions (which was also applied in developmental work on infants and children).

> Contrast gain control is related to 'automatic gain control' - a electrical engineering concept that arose in the early 20th century. The concept of modulating the input (rather than the output) of the system became established in the 1930s and 40s (for example in WW2 radar systems). The specific idea of computing that modulation as a function of the local average signal power seems to have arisen around this time and was refined over the 50s and 60s (for example, in 'sidechain compression'). In the domain of vision, an early example of 'contrast gain control' (scaling the input - in this case luminance) is found in Sperling and Sondhi 1968.

> The first formalization of the concept in the domain of spatial vision that included the critical concepts of an early, divisive gain control signal computed from some additional estimate of local contrast seems to appear in a seminal paper by Shapley and Victor, (1978). Application to contrast adaptation was demonstrated by Ohzawa, Sclar, Freeman, (1982, 1985) with a demonstration of instantaneous gain control in the human SSVEP by Bobak et al in 1988. A synthesis that included both adaptation and masking was proposed by Heeger 1992, while Foley 1994 showed an application to human psychophysical data. We have now cited these more foundational papers. We would be happy to include the relevant references to the non-linear biophysical model used in adult and developmental work if the reviewer can point us to the appropriate literature.

5. The description of Figure 3 and the nonlinear operations performed are unclear. They state in the text that Equation 2 (“hyperbolic ratio function”) is used as a nonlinear transducer, but the figure legend states that a squaring operator was used for frequency doubling. A squaring operation would only produce a sinusoid of double the frequency of the input. The waveform shown in the left panel appears to be a full-wave rectified signal, which would contain theoretically an infinite number of higher harmonics. The text also states that “This results from the distortion of the input sine waves at high contrast due to a combination of the full-wave rectification and saturating non-linearity.” When was full-wave rectification performed?

> We see that we went too fast here and the description of the signal processing was confusing. We now break out the two stages into (Figure 3) full wave rectification only (to model the contributions of on and off cells) - explicitly pointing out that this produces only even harmonics but does not alter the linearity of the CRF. We then show (Figure 4) the effect of passing this full wave rectified signal through a hyperbolic ratio function and illustrate the resulting saturation of the CRF (Figure 4). We think that breaking the model out in this way makes the contribution of each part clearer and allows the reader to play with the code for separate stages. The input contrast is squared in Figure 4 simply because the exponent we choose in the saturating nonlinearity is '2' but this is a parameter that is often fitted rather than fixed and the reader is encouraged to experiment with the exponent to examine its effect on the resulting amplitude spectrum.

> While there are, technically, infinite even harmonics in both Figure 3 and Figure 4, they decrease with frequency and we plot only three for reasons of space.

@@

6. The authors use the term “frequency tagging” which has been used in the literature to refer to stimulating different regions of the stimulus field or fellow eyes with different sinusoidal frequencies and measuring the intermodulation (sum and difference) frequency responses that result from nonlinear interactions. They do not, however, cite the foundational work in this field.

> As far as we are aware, the term refers to stimulating some *aspect* of the visual scene (a position or a feature) - not necessarily just a spatial region. There need not be more than one region (it is interesting, for example, to frequency tag a single feature or region and measure the effect of attending towards or away from it) and frequency tagged components can exist in the same location (for example, see the work on contrast normalization described above where the 'probe' and 'mask' have different tags and we explicitly examine the intermodulation terms). In fact much of this paper is concerned with discussing precisely these intermodulation terms and their origins. However, intermodulation is also not a 'necessary' part of a frequency tagging experiment: beautiful early work by Morgan et al (PNAS 1996) uses frequency tagging over space to examine attentional modulation purely in terms of the amplification or suppression of the F1 and F2 components. The tag need not even be visible - there is a fashion at the moment for 'rapid invisible frequency tagging' (RIFT) using flicker close to or beyond the CFF limit. In this sense, steady state paradigms are almost by definition frequency tagging paradigms because at least one feature or position is 'tagged'. In terms of using multiple input frequencies to characterise interactions between visual features, we think that Regan and Regan may have had the most comprehensive and some of the earliest descriptions of the method and we have credited them extensively throughout the paper. We are also happy to consider additional works if the reviewer can suggest them.

@@

7. There is a large body of work that used ssVEPs to investigate contrast processing developmentally (infants through adulthood) and in various disease states, which is mostly missed in this review article.

> We touch on both these areas but including them in detail would require us to extend what is already becoming a very long paper.

8. Figures should be labeled.

> It is possible that our text compilation stage failed to auto-generate the figure labels - we apologise if this is the case. The current version seems to be working correctly (in that each Figure has a label saying 'Figure 1', 'Figure 2' etc which is referenced in the main text).

@@

Thus, the stated intention of this manuscript (“Here we describe how SSVEPs have been used to study visual contrast over the past 70 years”) fails to be met. Only select articles from a few sources are covered, and therefore, the richness and value of this technique is lost.

> The original document had well over 100 references and the current bibliography now contains 156 entries. Although we have tried to be fair, we have of necessity 'selected' from the literature and have doubtless omitted many important foundational works. We are happy to take suggestions if parts of the field have been misrepresented.

> The original purpose of the document (now long eclipsed) was to serve as a convenient primer on the history of the technique and an interactive visualisation of key concepts implemented in Python and R using a novel 'executable paper' format. We think that getting this balance right means that we cannot include a comprehensive history of the entire field - in effect we must continue to be 'selective' and we hope that the relatively few papers we have cited here will serve as useful signposts to the richer literature.

### 2024-12-03 review.qmd
Minor typo edits, and hopefully fixed autobuild

@@

```
```{python fullwaverectification}

@@

```
```{python transducerout}

@@

Neurons in the visual areas of the brain are primarily responsive to changes in cone photoreceptor activations across time and space. This property, referred to as 'contrast', sets the fundamental limits of our visual abilities, which remain steady over a remarkably wide range of environmental light levels. The human response to contrast can be studied using many different techniques. Early work used psychophysical methods to measure contrast sensitivity [@Campbell1965, @schadeOpticalPhotoelectricAnalog1956], defined as the inverse of the lowest contrast that can be reliably detected. But neural responses can also be measured more directly using techniques such as functional magnetic resonance imaging (fMRI), magnetoencephalography (MEG), and electroencephalography (EEG). Here we will describe how an EEG method known as the steady state visually evoked potential (SSVEP) technique has contributed to our understanding of human contrast processing in health, disease and throughout development.

@@

Two common stimulus variants involve sinusoidal on-off (or 'appearance/disappearance') flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can excite independent populations of on- and off-cells in the visual system once per cycle. For spatial patterns, the contributions of individual cortical cells to these excitations are thought to sum at the scalp and generate an average of their onset responses on each cycle. For very low spatial frequencies (including zero) resposes at the scalp will be dominated by a single on- or off- cell type on each half cycle.

@@

By comparison, counterphase flickering patterns generate two essentially identical transients per cycle and therefore does not produce a response at 1F, only even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker (either on/off or contrast reversig) tends to produce additional spectral harmonics compared to sine-wave flicker.

@@

Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over space ('spatial contrast') or time ('temporal contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast, and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well [@reganEvokedPotentialsSpecific1973, @reganColourCodingPattern1975, @reganEffectStimulusColour1966a, @McKeefry1996; @Di-Russo2001] (see also [@Baseler1997]).

@@

These contrast definitions are appropriate both to photometric measures of stimulus contrast (for example, luminance; @Lennie1993) and also to definitions based on cone excitations [@Derrington1984; @MacLeod1979] which are more common in work on chromatic processing.

Although its mathematical definition is straightforward, the computations that underlie contrast processing in the brain have been the subject of intense research for many decades. The neural code for contrast, even in the earliest parts of visual cortex, is not simply a linear transform of the contrast at the retina - instead, contrast signals undergo a cascade of nonlinear processing stages that, broadly, attempt to normalise the output relative to the spatiotemporal environment. This normalization, achieved through a computation called 'contrast gain control' [@ohzawaContrastGainControl1982, @ohzawaContrastGainControl1985, @bobakCorticalContrastGain1988a, @Foley1994; @Heeger1992; @Carandini2011] maximises the sensitivity of the visual system by making optimal use of neuronal bandwidth. As an example, a grating placed at the centre of a low-contrast background typically appears more intense than the same grating when superimposed on a high contrast background (see @fig-centresurround; note that the code used to produce all figures in this review is available in python and R at: https://github.com/wadelab/contrastReviewPaperVNS).

@@

Although the SSVEP does seem to reflect the activity of relatively well-tuned neuronal populations , changing some aspect of the stimulus may change the nature of those populations. For example, Campbell and Maffei [@Campbell1970] noted that densely-sampled measurements of contrast responses can reveal the presence of two, qualitatively different types of neurons that exhibit different log-linear contrast functions (see also [@Tyler1985, @Nakayama1982, @Norcia1989 and @Souza2007]). Modeling such two-limbed contrast response functions using a single sigmoidal function is therefore an approximation.

@@

An additional complexity is introduced by the fact that the EEG is an average population response and that the visual system contains many different types of neurons. As a stimulus changes in, for example, contrast it may selectively activate qualitatively different neuronal populations. This type

To  understand these processes better, we will show how sinusoidal input signals might be processed by the visual system to produce SSVEPs. @fig-linear illustrates how sine waves of different contrasts are processed in a linear system. The first panel shows the input sine wave, which would be used to modulate stimulus amplitude over time. Notice that there are five peaks in the waveform during the one second sample, so the stimulation frequency is 5Hz (F1). The second panel shows Fourier transform of the waveform, which contains a substantial peak at this frequency. If we  change the stimulus contrast (i.e. the amplitude of the waveform), the amplitude of the F1 component increases linearly with contrast (right panel).

@@

First, we consider the fact that neurons in the early visual system code either positive or negative contrast. The responses of each different type of neuron is an increase in firing rate to the increasing levels of the prefered contrast polarity and these neurons typically exhibit zero (or baseline) response to the non-prefered polarity. We can model the combined responses of these two cell populations to a time varying contrast signal by full-wave-rectification. The effect of this computation is, effectively, to double the input freuquency in the population measurement and to introduce additional higher (even) harmonics due to the discontinuity at the contrast reversals. The effect of full-wave rectification on the SSVEP signal is shown in @fig-fullwaverectification

@@

Neurons in the visual system typically do not exhibit this perfect response linearity. Instead, neuronal response are well-modeled by some form of saturating non-linearity: as the contrast increases, the response of the neuron increases but the response increase per unit contrast is reduced at the high end of the contrast range. This saturating nonlinearity is often modeled as a hyperbolic ratio function of the form

@@

![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input wavefore (left panel) is frequency doubled by the full-wave rectification (see @fig-fullwaverectification), and now has energy at the second harmonic (10Hz) and higher even harmonics (middle panel). Passing the signal through a second nonlinearity (the hyperbolic ratio function) causes the contrast response function at the second harmonic (2F) to become nonlinear and it and saturates (or even 'supersaturates') at high contrasts (right panel).](){#fig-transducer}
:::
::: {.content-visible unless-format="docx"}
![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input wavefore (left panel) is frequency doubled by the full-wave rectification (see @fig-fullwaverectification), and now has energy at the second harmonic (10Hz) and higher even harmonics (middle panel). Passing the signal through a second nonlinearity (the hyperbolic ratio function) causes the contrast response function at the second harmonic (2F) to become nonlinear and it and saturates (or even 'supersaturates') at high contrasts (right panel).](Figures/4_transducer.pdf){#fig-transducer}
:::

To illustrate the effect of contrast gain control [@ohzawaContrastGainControl1982, @Heeger1992], we next include a second component (a 'mask') that contributes to the gain pool of the first (the 'target'). The mask stimulus will suppress the response at the target frequency, reducing its amplitude. The suppression is reciprocal - activity at the mask frequency is also reduced by the presence of the target, in a contrast-dependent manner [@Regan1988, @brownDevelopmentRivalryDichoptic1999, @Candy2001, @Busse2009]. At a single pair of (matched) contrast levels, we see a complex pattern of intermodulation terms in the transduced waveforms (left panel of @fig-twoinputs) and in the Fourier spectrum (middle panel of @fig-twoinputs). The two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms ---  is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan1988].

@@

where $C_\mathrm{mask}$ reflects the contrast of the mask component at a distinct frequency from that of the target. The effect of this extra term is to reduce the target response (see right panel of @fig-twoinputs). For a linear contrast axis, the contrast response function becomes shallower, whereas on a logarithmic contrast axis it maintains its steepness and shifts to the right. Suppressive effects of this kind have been obtained using SSVEP with a variety of different types of mask, including orthogonal overlaid masks, surround masks, and dichoptic masks [@Burr1987; @Ross1991; @Candy2001; @Busse2009; @Salelkar2020; @Cunningham2017; for a meta-analysis see @Baker2021].

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells [@Kuffler1957; @Alpern1971], and cells with similar properties in the lateral geniculate nucleus (LGN) [@Levitt2001] and cortex [@Movshon1975; @Hubel1959], will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion [@McKeefry1996]. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position [@Hubel1962]. Both the amplitude and the phase of the SSVEP response to a single contrast-reversing sine-wave grating therefore contains information about nonlinear computations performed across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain signal arising from non-linear interactions is intriguing. Presumably, the signal measured from early visual cortex is the result of a cascade of nonlinear retinal and cortical operations up to that point. It therefore contains a 'signature' or 'fingerprint' of the computational nature, order and parameters of those operations - including the shape of the transducer functions and the computations involved in signal combination. In principle, that information could be recovered from the SSVEP signal - a possibility recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is challenging, and a full system identification may require the use of multiple different input frequencies [@boydMeasuringVolterraKernels1983], careful parametric variation of the input stimuli does allow us to fit models of early visual visual processing (e.g. [@Zemon2006Dynamic]) and, by incorporating mathematical models of gain control mechanisms with either instantaneous, or temporally-integrated gain control, it is possible to model gain control using SSVEP data. For example, [@Tsai2012] demonstrated that a relatively simple gain control model gave a good account of the pattern of intermodulation responses produced by two overlaid patterns flickering at different frequencies. This was achieved by passing full stimulus waveforms through the transducer nonlinearity, and calculating the Fourier spectrum of the model output. More recent work has used this approach to study binocular integration in both normal subjects and amblyopes[@Hou-houchuan2021,@Hou2020]. Our more recent work on signal combination across eyes and space similarly demonstrated close correspondence between the predictions of a computational model and empirical data in humans [@Baker2017]. More detailed modelling of intracortical recordings [@Groen2022] has revealed details of the timecourse of gain control effects, specifically that normalization is delayed slightly relative to the initial visual response.

Similar changes to the contrast response function might also be obtained using adaptation paradigms, in which the visual system is exposed to high contrast stimuli for long durations. Psychophysically, adaptation increases detection thresholds, but has little effect on contrast discrimination performance [@Ross1993], much like pattern masks [@Foley1994]. Although SSVEP adaptation effects show strong tuning for orientation [@Vergeer2018; @Campbell1970] and spatial frequency [@Mecacci1976], there appear to be very measurements of the full SSVEP contrast response function before and after an extended period of adaptation. One exception is a study by Bach et al [@bachGreenleeBuhler1988] that examined the effects of adaptation on SSVEP responses who found a paradoxical increase in the 1F SSVEP amplitude following prolonged exposure to high-contrast stimuli across much of the probe contrast range. They explain this as being due to the presence of a nonlinear transducer function that saturates at high probe contrast driving the response to higher harmonic frequencies. Adaptation reduces the apparent contrast of the input, leading to less distortion, a more sinusoidal output and a concentration of the frequency-tagged response at lower harmonics. This explanation is conceptually related to the 'supersaturation' effect that we observe in Figure 4.

@@

Because SSVEP responses at detection threshold are very small, estimating a threshold is achieved by measuring the contrast response function at relatively high levels (in the linear part of the log-contrast response function prior to saturation), and extrapolating back along the function (either contrast vs response measured at a constant spatial frequency or spatial frequency vs response at a constant contrast level) to estimate its intercept with the x-axis [@Campbell1970] (see @fig-sweepvep). This contrast level has been shown to correspond approximately with psychophysically measured detection thresholds [@campbellVisualEvokedPotential1972, @Norcia1986]. One factor affecting the accuracy of this technique is the presence of noise in the EEG signal which will increase the apparent amplitude of the SSVEP responses across the measurement range. The effect of this noise is particulary apparent for recordings with low SNR or where the slope of the CRF is shallow (and therefore where small changes in the vertical offset of the fit result in large changes in the X-axis intercept). Accurate estimates of thresholds therefore require the experimenter to model this noise and compensate for its effects - as detailed in [@Norcia1989].

@@

The SSVEP technique has also been used to study the development of the contrast gain control mechanisms described in the previous section. Although response- [@Morrone1986] or contrast gain control [@Candy2001; ; @Skoczenski1998] is measurable in infants as young as six weeks old, its development appears to be slower, with adult levels being reached at approximately 11 years [@Pei2017].

>>>>

if not os.path.exists('Figures/'):
  os.makedirs('Figures/')
```

@@

```{python fullwaverectification}

@@

```{python transducerout}

@@

Neurons in the visual areas of the brain are primarily responsive to changes in cone photoreceptor activations across time and space. This property, referred to as 'contrast', sets the fundamental limits of our visual abilities, which remain steady over a remarkably wide range of environmental light levels. The human response to contrast can be studied using many different techniques. Early work used psychophysical methods to measure contrast sensitivity [@Campbell1965; @schadeOpticalPhotoelectricAnalog1956], defined as the inverse of the lowest contrast that can be reliably detected. But neural responses can also be measured more directly using techniques such as functional magnetic resonance imaging (fMRI), magnetoencephalography (MEG), and electroencephalography (EEG). Here we will describe how an EEG method known as the steady state visually evoked potential (SSVEP) technique has contributed to our understanding of human contrast processing in health, disease and throughout development.

@@

Two common stimulus variants involve sinusoidal on-off (or 'appearance/disappearance') flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can excite independent populations of on- and off-cells in the visual system once per cycle. For spatial patterns, the contributions of individual cortical cells to these excitations are thought to sum at the scalp and generate an average of their onset responses on each cycle. For very low spatial frequencies (including zero), resposes at the scalp will be dominated by a single on- or off- cell type on each half cycle.

@@

By comparison, counterphase flickering patterns generate two essentially identical transients per cycle and therefore do not produce a response at 1F, only even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker (either on/off or contrast reversing) tends to produce additional spectral harmonics compared to sine-wave flicker.

@@

Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over space ('spatial contrast') or time ('temporal contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast, and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well [@reganEvokedPotentialsSpecific1973; @reganColourCodingPattern1975; @reganEffectStimulusColour1966a; @McKeefry1996; @Di-Russo2001] (see also [@Baseler1997]).

@@

These contrast definitions are appropriate both to photometric measures of stimulus contrast [for example, luminance; @Lennie1993] and also to definitions based on cone excitations [@Derrington1984; @MacLeod1979] which are more common in work on chromatic processing.

Although its mathematical definition is straightforward, the computations that underlie contrast processing in the brain have been the subject of intense research for many decades. The neural code for contrast, even in the earliest parts of visual cortex, is not simply a linear transform of the contrast at the retina - instead, contrast signals undergo a cascade of nonlinear processing stages that, broadly, attempt to normalise the output relative to the spatiotemporal environment. This normalization, achieved through a computation called 'contrast gain control' [@ohzawaContrastGainControl1982; @ohzawaContrastGainControl1985; @bobakCorticalContrastGain1988a; @Foley1994; @Heeger1992; @Carandini2011] maximises the sensitivity of the visual system by making optimal use of neuronal bandwidth. As an example, a grating placed at the centre of a low-contrast background typically appears more intense than the same grating when superimposed on a high contrast background (see @fig-centresurround; note that the code used to produce all figures in this review is available in python and R at: https://github.com/wadelab/contrastReviewPaperVNS).

@@

Although the SSVEP does seem to reflect the activity of relatively well-tuned neuronal populations, changing some aspect of the stimulus may change the nature of those populations. For example, @Campbell1970 noted that densely-sampled measurements of contrast responses can reveal the presence of two qualitatively different types of neurons that exhibit different log-linear contrast functions [see also @Tyler1985; @Nakayama1982; @Norcia1989; @Souza2007]. Modeling such two-limbed contrast response functions using a single sigmoidal function is therefore an approximation.

@@

An additional complexity is introduced by the fact that the EEG is an average population response and that the visual system contains many different types of neurons. As a stimulus changes in (for example) contrast, it may selectively activate qualitatively different neuronal populations.

To  understand these processes better, we will show how sinusoidal input signals might be processed by the visual system to produce SSVEPs. @fig-linear illustrates how sine waves of different contrasts are processed in a linear system. The first panel shows the input sine wave, which would be used to modulate stimulus amplitude over time. Notice that there are five peaks in the waveform during the one second sample, so the stimulation frequency is 5Hz (F1). The second panel shows the Fourier transform of the waveform, which contains a substantial peak at this frequency. If we change the stimulus contrast (i.e. the amplitude of the waveform), the amplitude of the F1 component increases linearly with contrast (right panel).

@@

First, we consider the fact that neurons in the early visual system code either positive or negative contrast. The response of each different type of neuron is an increase in firing rate to the increasing levels of the preferred contrast polarity, and these neurons typically exhibit zero (or baseline) response to the non-preferred polarity. We can model the combined responses of these two cell populations to a time varying contrast signal by full-wave-rectification. The effect of this computation is, effectively, to double the input frequency in the population measurement and to introduce additional higher (even) harmonics due to the discontinuity at the contrast reversals. The effect of full-wave rectification on the SSVEP signal is shown in @fig-fullwaverectification

@@

Neurons in the visual system typically do not exhibit this perfect response linearity. Instead, neuronal response are well-modeled by some form of saturating non-linearity: as the contrast increases, the response of the neuron increases but the response increase per unit contrast is reduced at the high end of the contrast range. This saturating nonlinearity is often modeled as a hyperbolic ratio function of the form:

@@

![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input waveform (left panel) is frequency doubled by the full-wave rectification (see @fig-fullwaverectification), and now has energy at the second harmonic (10Hz) and higher even harmonics (middle panel). Passing the signal through a second nonlinearity (the hyperbolic ratio function) causes the contrast response function at the second harmonic (2F) to become nonlinear and it and saturates (or even 'supersaturates') at high contrasts (right panel).](){#fig-transducer}
:::
::: {.content-visible unless-format="docx"}
![Illustration of how a sinusoidal input signal is affected by a nonlinear transducer. The input waveform (left panel) is frequency doubled by the full-wave rectification (see @fig-fullwaverectification), and now has energy at the second harmonic (10Hz) and higher even harmonics (middle panel). Passing the signal through a second nonlinearity (the hyperbolic ratio function) causes the contrast response function at the second harmonic (2F) to become nonlinear and it and saturates (or even 'supersaturates') at high contrasts (right panel).](Figures/4_transducer.pdf){#fig-transducer}
:::

To illustrate the effect of contrast gain control [@ohzawaContrastGainControl1982; @Heeger1992], we next include a second component (a 'mask') that contributes to the gain pool of the first (the 'target'). The mask stimulus will suppress the response at the target frequency, reducing its amplitude. The suppression is reciprocal - activity at the mask frequency is also reduced by the presence of the target, in a contrast-dependent manner [@Regan1988; @brownDevelopmentRivalryDichoptic1999; @Candy2001; @Busse2009]. At a single pair of (matched) contrast levels, we see a complex pattern of intermodulation terms in the transduced waveforms (left panel of @fig-twoinputs) and in the Fourier spectrum (middle panel of @fig-twoinputs). The two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms ---  is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan1988].

@@

where $C_\mathrm{mask}$ reflects the contrast of the mask component at a frequency distinct from that of the target. The effect of this extra term is to reduce the target response (see right panel of @fig-twoinputs). For a linear contrast axis, the contrast response function becomes shallower, whereas on a logarithmic contrast axis it maintains its steepness and shifts to the right. Suppressive effects of this kind have been obtained using SSVEP with a variety of different types of mask, including orthogonal overlaid masks, surround masks, and dichoptic masks [@Burr1987; @Ross1991; @Candy2001; @Busse2009; @Salelkar2020; @Cunningham2017; for a meta-analysis see @Baker2021].

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells [@Kuffler1957; @Alpern1971], and cells with similar properties in the lateral geniculate nucleus (LGN) [@Levitt2001] and cortex [@Movshon1975; @Hubel1959], will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion [@McKeefry1996]. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position [@Hubel1962]. Both the amplitude and the phase of the SSVEP response to a single contrast-reversing sine-wave grating therefore contain information about nonlinear computations performed across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain signal arising from non-linear interactions is intriguing. Presumably, the signal measured from early visual cortex is the result of a cascade of nonlinear retinal and cortical operations up to that point. It therefore contains a 'signature' or 'fingerprint' of the computational nature, order and parameters of those operations - including the shape of the transducer functions and the computations involved in signal combination. In principle, that information could be recovered from the SSVEP signal - a possibility recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along the entire processing pathway is challenging, and a full system identification may require the use of multiple different input frequencies [@boydMeasuringVolterraKernels1983], careful parametric variation of the input stimuli does allow us to fit models of early visual visual processing [e.g. @Zemon2006Dynamic] and, by incorporating mathematical models of gain control mechanisms with either instantaneous, or temporally-integrated gain control, it is possible to model gain control using SSVEP data. For example, @Tsai2012 demonstrated that a relatively simple gain control model gave a good account of the pattern of intermodulation responses produced by two overlaid patterns flickering at different frequencies. This was achieved by passing full stimulus waveforms through the transducer nonlinearity, and calculating the Fourier spectrum of the model output. More recent work has used this approach to study binocular integration in both normal subjects and amblyopes [@Hou-houchuan2021; @Hou2020]. Our more recent work on signal combination across eyes and space similarly demonstrated close correspondence between the predictions of a computational model and empirical data in humans [@Baker2017]. More detailed modelling of intracortical recordings [@Groen2022] has revealed details of the timecourse of gain control effects, specifically that normalization is delayed slightly relative to the initial visual response.

Similar changes to the contrast response function might also be obtained using adaptation paradigms, in which the visual system is exposed to high contrast stimuli for long durations. Psychophysically, adaptation increases detection thresholds, but has little effect on contrast discrimination performance [@Ross1993], much like pattern masks [@Foley1994]. Although SSVEP adaptation effects show strong tuning for orientation [@Vergeer2018; @Campbell1970] and spatial frequency [@Mecacci1976], there appear to be very few measurements of the full SSVEP contrast response function before and after an extended period of adaptation. One exception is a study by @bachGreenleeBuhler1988 that examined the effects of adaptation on SSVEP responses, and found a paradoxical increase in the 1F SSVEP amplitude following prolonged exposure to high-contrast stimuli across much of the probe contrast range. They explain this as being due to the presence of a nonlinear transducer function that saturates at high probe contrasts, driving the response to higher harmonic frequencies. Adaptation reduces the apparent contrast of the input, leading to less distortion, a more sinusoidal output and a concentration of the frequency-tagged response at lower harmonics. This explanation is conceptually related to the 'supersaturation' effect that we observed in Figure 4.

@@

Because SSVEP responses at detection threshold are very small, estimating a threshold is achieved by measuring the contrast response function at relatively high levels (in the linear part of the log-contrast response function prior to saturation), and extrapolating back along the function (either contrast vs response measured at a constant spatial frequency or spatial frequency vs response at a constant contrast level) to estimate its intercept with the x-axis [@Campbell1970] (see @fig-sweepvep). This contrast level has been shown to correspond approximately with psychophysically measured detection thresholds [@campbellVisualEvokedPotential1972; @Norcia1986]. One factor affecting the accuracy of this technique is the presence of noise in the EEG signal which will increase the apparent amplitude of the SSVEP responses across the measurement range. The effect of this noise is particulary apparent for recordings with low SNR or where the slope of the CRF is shallow (and therefore where small changes in the vertical offset of the fit result in large changes in the X-axis intercept). Accurate estimates of thresholds therefore require the experimenter to model this noise and compensate for its effects - as detailed in [@Norcia1989].

@@

The SSVEP technique has also been used to study the development of the contrast gain control mechanisms described in the previous section. Although response- [@Morrone1986] or contrast gain control [@Candy2001; @Skoczenski1998] is measurable in infants as young as six weeks old, its development appears to be slower, with adult levels being reached at approximately 11 years [@Pei2017].
